---
title: "Lectures 5 & 6 — Multi-State Life Tables"
author: "Dr Belen Zapata-Diomedi, IMS Epidemiology Unit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    theme: flatly
    highlight-style: tango
    code-fold: show
    code-tools: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
  fig-width: 10
  fig-height: 4
---

# Data and Functions {#sec-data}

Here we load the libraries, source the helper functions, and read all required data.

```{r}
#| label: libraries-and-functions

library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)

# ── Helper functions (RunLifeTable, assignPA_RRs, calculate_pif) ───────────────
source("functions.R")

# ── Output directory ───────────────────────────────────────────────────────────
dir.create("outputs/lecture_5_6", recursive = TRUE, showWarnings = FALSE)
```

## MSLT data (Colombia) {#sec-mslt-data}

The MSLT data file contains one row per age (0–100) and sex, with all-cause mortality rates (`mx`), disability weights (`pyld_rate`), population size, and disease-specific parameters for Colombia. All rates are derived from GBD 2023
Colombia estimates.

```{r}
#| label: load-mslt

mslt_general <- read_csv("data/lecture_5_6/mslt_colombia.csv") #(CHANGE for final file)

glimpse(mslt_general)
```

## Bogotá population data {#sec-bogota-data}

This file contains the Bogotá study population with marginal MET-hours per week drawn from sex-specific log-normal distributions, plus a counterfactual scenario (`mmet_scen = mmet_ref + 2`). It was prepared in Lecture 3.

```{r}
#| label: load-bogota

bogota_mmets <- read_csv("data/base_pop_bogota_ph.csv")

bogota_mmets %>%
  pivot_longer(
    cols      = c(mmet_ref, mmet_scen),
    names_to  = "scenario",
    values_to = "mmet"
  ) %>%
  mutate(scenario = ifelse(scenario == "mmet_ref",
                           "Reference", "Scenario (+2 mMETs)")) %>%
  group_by(sex, scenario) %>%
  summarise(
    n      = n(),
    min    = round(min(mmet,             na.rm = TRUE), 2),
    q1     = round(quantile(mmet, 0.25,  na.rm = TRUE), 2),
    median = round(median(mmet,          na.rm = TRUE), 2),
    mean   = round(mean(mmet,            na.rm = TRUE), 2),
    q3     = round(quantile(mmet, 0.75,  na.rm = TRUE), 2),
    max    = round(max(mmet,             na.rm = TRUE), 2),
    .groups = "drop"
  )
```

---

# Life Table Structure and Variables {#sec-structure}

A **cohort life table** tracks a group of people born (or entering an age) at the same time and follows them as they age and die. Each row represents one year of age.

| Variable    | Meaning |
|-------------|---------|
| `mx`        | All-cause mortality *rate* at age x (deaths per person-year) |
| `qx`        | Probability of dying between age x and x+1 |
| `lx`        | Number of survivors at age x |
| `dx`        | Number who die between age x and x+1 |
| `Lx`        | Person-years lived by the cohort between age x and x+1 |
| `ex`        | Remaining life expectancy at age x |
| `pyld_rate` | All-cause disability weight (proportion of health lost) |
| `Lwx`       | Health-adjusted person-years lived (Lx × (1 − pyld_rate)) |
| `ewx`       | Health-adjusted life expectancy at age x |

The key relationships are:

$$q_x = 1 - e^{-m_x}$$

$$L_x = \frac{l_x + l_{x+1}}{2}$$

$$e_x = \frac{\sum_{t=x}^{100} L_t}{l_x}$$

## Reference life table (no trend) {#sec-reference}

We work with **50–54-year-old females** (mid-age = 52) throughout. Change `in_sex` and `in_mid_age` to explore other cohorts.

Available mid-ages: 17, 22, 27, 32, 37, 42, 47, **52**, 57, 62, 67, 72, 77, 82, 87, 92, 97

```{r}
#| label: reference-life-table

lt_ref <- RunLifeTable(
  in_data    = mslt_general,
  in_sex     = "female",
  in_mid_age = 52,
  mx_trend   = NA,   # NA = no trend, use static mx
  scenario   = 0     # 0 = no PIF-based reduction
) %>%
  mutate(label = "Reference (no trend)")

write_csv(lt_ref, "outputs/lecture_5_6/life_table_reference.csv")

lt_ref %>%
  select(age, mx, qx, lx, dx, Lx, ex, Lwx, ewx) %>%
  head(10)
```

```{r}
#| label: inspect-lt

cat(sprintf("Remaining life expectancy at age 52 (female): %.2f years\n",
            lt_ref$ex[1]))
cat(sprintf("Health-adjusted life expectancy at age 52:    %.2f years\n",
            lt_ref$ewx[1]))
cat(sprintf("\nTotal person-years lived by cohort (sum Lx):  %s\n",
            format(round(sum(lt_ref$Lx)), big.mark = ",")))
cat(sprintf("Total health-adjusted life years  (sum Lwx):  %s\n",
            format(round(sum(lt_ref$Lwx)), big.mark = ",")))
```

```{r}
#| label: plot-reference
#| fig-cap: "Reference life table for females aged 52+. Left: all-cause mortality rate rising with age. Right: cohort survivors declining over time."

p1 <- ggplot(lt_ref, aes(x = age, y = mx)) +
  geom_line(colour = "#2563EB", linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "All-cause mortality rate — females aged 52+",
       x = "Age", y = "Mortality rate (mx)") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11))

p2 <- ggplot(lt_ref, aes(x = age, y = lx / 1e6)) +
  geom_line(colour = "#0D9488", linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Survivors — females aged 52+",
       x = "Age", y = "Survivors (millions)") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## What if mortality improves over time? {#sec-trend}

The reference life table above uses **static** mortality rates — the same `mx` at every age, regardless of which calendar year the cohort reaches that age. This is a simplification. In reality, mortality rates change over time as healthcare, living standards, and disease management improve (or worsen).

### How the trend is modelled

The `mx_trend` argument applies an **annual percentage change** that **compounds** across simulation years. Because each row of the life table represents the cohort one year older — and therefore one calendar year further into the future — the multiplier accumulates:

$$m_x^{\text{projected}} = m_x \times \left(1 - \frac{\text{trend}}{100}\right)^{t-1}$$

where $t = 1$ at cohort entry (no change), $t = 2$ one year later, and so on.

::: {.callout-note}
## Why compound, not linear?
A linear reduction of 0.5% per year would eventually drive `mx` negative, which is impossible. Compounding means each year's rate is 0.5% lower than the *already-improved* previous year — both mathematically correct and epidemiologically realistic.
:::

The table below shows how a 0.5% annual reduction accumulates for a starting `mx = 0.01`:

| Simulation year | Age | Multiplier | Projected mx |
|---:|---:|---:|---:|
| 1 | 52 | $(0.995)^0 = 1.0000$ | 0.01000 |
| 2 | 53 | $(0.995)^1 = 0.9950$ | 0.00995 |
| 5 | 56 | $(0.995)^4 = 0.9802$ | 0.00980 |
| 10 | 61 | $(0.995)^9 = 0.9559$ | 0.00956 |
| 20 | 71 | $(0.995)^{19} = 0.9085$ | 0.00909 |
| 30 | 81 | $(0.995)^{29} = 0.8636$ | 0.00864 |
| 49 | 100 | $(0.995)^{48} = 0.7851$ | 0.00785 |

By the terminal age the mortality rate is ~21% lower than at entry, reflecting nearly 50 years of accumulated improvement.

### Running the life table with a 0.5% annual trend

```{r}
#| label: lt-with-trend

lt_trend <- RunLifeTable(
  in_data    = mslt_general,
  in_sex     = "female",
  in_mid_age = 52,
  mx_trend   = 0.5,   # 0.5% annual reduction in mx, compounding each year
  scenario   = 0
) %>%
  mutate(label = "With trend (−0.5%/yr)")

write_csv(lt_trend, "outputs/lecture_5_6/life_table_trend.csv")
```

### Comparing reference vs trend

```{r}
#| label: compare-trend

gain_ly_trend  <- sum(lt_trend$Lx)  - sum(lt_ref$Lx)
gain_hly_trend <- sum(lt_trend$Lwx) - sum(lt_ref$Lwx)
gain_ex_trend  <- (lt_trend$ex[1]   - lt_ref$ex[1]) * 365

cat(sprintf("Difference in person-years lived (Lx):           %s\n",
            format(round(gain_ly_trend),  big.mark = ",")))
cat(sprintf("Difference in health-adjusted life years (Lwx):  %s\n",
            format(round(gain_hly_trend), big.mark = ",")))
cat(sprintf("Difference in life expectancy:                   %.1f days (%.2f years)\n",
            gain_ex_trend, gain_ex_trend / 365))
```

```{r}
#| label: plot-trend-comparison
#| fig-cap: "Effect of a 0.5% annual mortality reduction on the mortality rate trajectory (left) and cumulative person-years lived (right). The gap between curves widens with age as the compounding effect accumulates."

lt_compare_trend <- bind_rows(lt_ref, lt_trend)

p_mx <- ggplot(lt_compare_trend, aes(x = age, y = mx, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"    = "#2563EB",
                                 "With trend (−0.5%/yr)"   = "#7C3AED")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Mortality rate: reference vs 0.5% annual trend",
       x = "Age", y = "Mortality rate (mx)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

p_lx <- ggplot(lt_compare_trend, aes(x = age, y = Lx / 1e6, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"    = "#2563EB",
                                 "With trend (−0.5%/yr)"   = "#7C3AED")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Person-years lived: reference vs 0.5% annual trend",
       x = "Age", y = "Person-years (millions)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

gridExtra::grid.arrange(p_mx, p_lx, ncol = 2)
```

```{r}
#| label: summary-table-trend
#| echo: false

tibble(
  Scenario            = c("Reference (no trend)",
                           "With trend (−0.5%/yr)",
                           "Difference"),
  `Life expectancy`   = c(round(lt_ref$ex[1],   2),
                           round(lt_trend$ex[1], 2),
                           round(lt_trend$ex[1] - lt_ref$ex[1], 2)),
  `HA life expectancy`= c(round(lt_ref$ewx[1],   2),
                           round(lt_trend$ewx[1], 2),
                           round(lt_trend$ewx[1] - lt_ref$ewx[1], 2)),
  `Sum Lx`            = c(format(round(sum(lt_ref$Lx)),   big.mark = ","),
                           format(round(sum(lt_trend$Lx)), big.mark = ","),
                           format(round(gain_ly_trend),    big.mark = ",")),
  `Sum Lwx`           = c(format(round(sum(lt_ref$Lwx)),   big.mark = ","),
                           format(round(sum(lt_trend$Lwx)), big.mark = ","),
                           format(round(gain_hly_trend),    big.mark = ","))
) |> knitr::kable()
```

::: {.callout-tip}
## Interpretation
A 0.5% annual compound reduction in all-cause mortality  adds approximately `r format(round(gain_ly_trend), big.mark = ",")` person-years to this cohort over their remaining lifetime and increases life expectancy by about `r round(gain_ex_trend)` days. The effect is larger than it may seem because compounding means the improvement accumulates silently over decades.
:::

::: {.callout-note}
## Extending this
`mx_trend` can also be **negative** to model worsening mortality (e.g. `mx_trend = -1` for a 1% annual increase). Any value can be stacked on top of a PIF-based `scenario` reduction — the two mechanisms are applied independently and in sequence.
:::

---

# Scenario 1 — Direct Mortality Reduction (5%) {#sec-scenario1}

**Question:** What is the gain in life years and life expectancy if all-cause mortality is reduced by 5% across all ages?

The `scenario` argument applies a one-off proportional reduction to `mx` at every age:

$$m_x^{\text{scenario}} = m_x \times (1 - \text{scenario})$$

This is distinct from `mx_trend`, which compounds year-on-year. The two can be combined: `scenario` captures a structural shift (e.g. from a policy intervention) while `mx_trend` captures background secular change.

```{r}
#| label: scenario-mortality-reduction

lt_sc5 <- RunLifeTable(
  in_data    = mslt_general,
  in_sex     = "female",
  in_mid_age = 52,
  mx_trend   = NA,
  scenario   = 0.05      # 5% one-off reduction at all ages
) %>%
  mutate(label = "5% mortality reduction")

write_csv(lt_sc5, "outputs/lecture_5_6/life_table_scenario_5pct.csv")
```

## Comparing reference vs scenario

```{r}
#| label: compare-scenarios

gain_ly  <- sum(lt_sc5$Lx)  - sum(lt_ref$Lx)
gain_hly <- sum(lt_sc5$Lwx) - sum(lt_ref$Lwx)
gain_ex  <- (lt_sc5$ex[1]   - lt_ref$ex[1]) * 365

cat(sprintf("Gain in person-years lived (Lx):           %s\n",
            format(round(gain_ly),  big.mark = ",")))
cat(sprintf("Gain in health-adjusted life years (Lwx):  %s\n",
            format(round(gain_hly), big.mark = ",")))
cat(sprintf("Gain in life expectancy:                   %.1f days (%.2f years)\n",
            gain_ex, gain_ex / 365))
```

```{r}
#| label: plot-scenario-comparison
#| fig-cap: "5% one-off mortality reduction vs reference. Unlike the trend, the gap between curves is constant across ages — the proportional reduction is uniform."

lt_compare <- bind_rows(lt_ref, lt_sc5)

p3 <- ggplot(lt_compare, aes(x = age, y = mx, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"   = "#2563EB",
                                 "5% mortality reduction" = "#DC2626")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Mortality rate: reference vs 5% reduction",
       x = "Age", y = "Mortality rate (mx)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

p4 <- ggplot(lt_compare, aes(x = age, y = Lx / 1e6, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"   = "#2563EB",
                                 "5% mortality reduction" = "#DC2626")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Person-years lived: reference vs 5% reduction",
       x = "Age", y = "Person-years (millions)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

gridExtra::grid.arrange(p3, p4, ncol = 2)
```

## Summary table — reference, trend, and scenario

```{r}
#| label: summary-table-scenario1
#| echo: false

tibble(
  Scenario            = c("Reference (no trend)",
                           "With trend (−0.5%/yr)",
                           "5% mortality reduction",
                           "Difference (trend vs ref)",
                           "Difference (5% sc vs ref)"),
  `Life expectancy`   = c(round(lt_ref$ex[1],   2),
                           round(lt_trend$ex[1], 2),
                           round(lt_sc5$ex[1],   2),
                           round(lt_trend$ex[1] - lt_ref$ex[1], 2),
                           round(lt_sc5$ex[1]   - lt_ref$ex[1], 2)),
  `HA life expectancy`= c(round(lt_ref$ewx[1],   2),
                           round(lt_trend$ewx[1], 2),
                           round(lt_sc5$ewx[1],   2),
                           round(lt_trend$ewx[1] - lt_ref$ewx[1], 2),
                           round(lt_sc5$ewx[1]   - lt_ref$ewx[1], 2)),
  `Sum Lx`            = c(format(round(sum(lt_ref$Lx)),   big.mark = ","),
                           format(round(sum(lt_trend$Lx)), big.mark = ","),
                           format(round(sum(lt_sc5$Lx)),   big.mark = ","),
                           format(round(gain_ly_trend),    big.mark = ","),
                           format(round(gain_ly),          big.mark = ","))
) |> knitr::kable()
```

::: {.callout-tip}
## Scenario vs trend — a key teaching point
Notice that a **5% one-off reduction** and a **0.5% annual trend** can produce similar life-year gains over a long cohort follow-up, yet they represent very different processes. The one-off reduction applies uniformly at all ages from the start; the trend starts small and grows, with the largest benefit arriving late in the cohort's life. When evaluating interventions, it matters whether the mechanism is a permanent structural shift or a gradual secular change.
:::

---

# Scenario 2 — Physical Activity Increase (Bogotá) {#sec-scenario2}

**Question:** What is the gain in life years from a population-level increase in physical activity (+2 mMET-hours/week) for the Bogotá population?

The pathway is:

$$\text{PA distribution} \xrightarrow{\text{dose-response}} \text{RR by individual} \xrightarrow{\text{PIF}} \Delta m_x \xrightarrow{\text{life table}} \Delta \text{life years}$$

## Step 1 — Assign individual-level RRs {#sec-rrs}

```{r}
#| label: compute-rrs

diseases <- "all-cause-mortality"

bogota_ref  <- assignPA_RRs(bogota_mmets, diseases = diseases,
                             mmet_col = "mmet_ref")
bogota_scen <- assignPA_RRs(bogota_mmets, diseases = diseases,
                             mmet_col = "mmet_scen")

# Sanity check — RR should be <= 1
cat("Reference RR summary:\n")
summary(bogota_ref$`rr_PHYSICAL_ACTIVITY_all-cause-mortality`)
cat("\nScenario RR summary (should be slightly lower):\n")
summary(bogota_scen$`rr_PHYSICAL_ACTIVITY_all-cause-mortality`)
```

## Step 2 — Calculate PIFs {#sec-pifs}

$$\text{PIF} = \frac{\sum \text{RR}(\text{ref}) - \sum \text{RR}(\text{scen})}{\sum \text{RR}(\text{ref})}$$

```{r}
#| label: compute-pifs

pif_results <- calculate_pif(pop_ref = bogota_ref, pop_scen = bogota_scen)
pifs        <- pif_results$pif_table

print(pifs, n = Inf)
```

## Step 3 — Extract PIF for the 50–54 female cohort (NEEDS fixing, PIFS are age specific) {#sec-extract-pif}

```{r}
#| label: extract-pif-for-cohort

pif_female_5054 <- pifs %>%
  filter(sex == "female", age_cat == "50-54") %>%
  pull(pif_pa)

cat(sprintf("PIF for all-cause mortality (female, 50–54): %.4f (%.2f%%)\n",
            pif_female_5054, pif_female_5054 * 100))
```

## Step 4 — Run life table with PIF-modified mortality {#sec-lt-pa}

$$m_x^{\text{PA scenario}} = m_x \times (1 - \text{PIF})$$

```{r}
#| label: life-table-pa-scenario

lt_pa <- RunLifeTable(
  in_data    = mslt_general,
  in_sex     = "female",
  in_mid_age = 52,
  mx_trend   = NA,
  scenario   = pif_female_5054
) %>%
  mutate(label = "PA scenario (+2 mMETs)")

write_csv(lt_pa, "outputs/lecture_5_6/life_table_pa_scenario.csv")
```

## Step 5 — Compare reference vs PA scenario {#sec-compare-pa}

```{r}
#| label: compare-pa-scenario

gain_ly_pa  <- sum(lt_pa$Lx)  - sum(lt_ref$Lx)
gain_hly_pa <- sum(lt_pa$Lwx) - sum(lt_ref$Lwx)
gain_ex_pa  <- (lt_pa$ex[1]   - lt_ref$ex[1]) * 365

cat(sprintf("Gain in person-years lived (Lx):          %s\n",
            format(round(gain_ly_pa),  big.mark = ",")))
cat(sprintf("Gain in health-adjusted life years (Lwx): %s\n",
            format(round(gain_hly_pa), big.mark = ",")))
cat(sprintf("Gain in life expectancy:                  %.1f days\n",
            gain_ex_pa))
```

```{r}
#| label: plot-pa-scenario
#| fig-cap: "PA scenario vs reference. The PIF reduces mx proportionally at all ages, similar in shape to the 5% one-off reduction but typically smaller in magnitude."

lt_compare_pa <- bind_rows(lt_ref, lt_pa)

p5 <- ggplot(lt_compare_pa, aes(x = age, y = mx, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"    = "#2563EB",
                                 "PA scenario (+2 mMETs)"  = "#D97706")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Mortality rate: reference vs PA scenario",
       x = "Age", y = "Mortality rate (mx)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

p6 <- ggplot(lt_compare_pa, aes(x = age, y = Lx / 1e6, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference (no trend)"    = "#2563EB",
                                 "PA scenario (+2 mMETs)"  = "#D97706")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10),
                     expand = c(0, 0)) +
  labs(title = "Person-years lived: reference vs PA scenario",
       x = "Age", y = "Person-years (millions)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text  = element_text(size = 11),
        axis.title = element_text(size = 11),
        legend.position = "bottom")

gridExtra::grid.arrange(p5, p6, ncol = 2)
```

## Summary table — all scenarios {#sec-summary}

```{r}
#| label: final-summary-table
#| echo: false

tibble(
  Scenario             = c("Reference (no trend)",
                            "With trend (−0.5%/yr)",
                            "5% mortality reduction",
                            "PA scenario (+2 mMETs)"),
  `mx_trend`           = c("NA", "0.5", "NA", "NA"),
  `scenario (PIF)`     = c("0", "0",
                            "5%",
                            paste0(round(pif_female_5054 * 100, 2), "%")),
  `Life expectancy`    = c(round(lt_ref$ex[1],   2),
                            round(lt_trend$ex[1], 2),
                            round(lt_sc5$ex[1],   2),
                            round(lt_pa$ex[1],    2)),
  `HA life expectancy` = c(round(lt_ref$ewx[1],   2),
                            round(lt_trend$ewx[1], 2),
                            round(lt_sc5$ewx[1],   2),
                            round(lt_pa$ewx[1],    2)),
  `Gain Lx`            = c("—",
                            format(round(gain_ly_trend), big.mark = ","),
                            format(round(gain_ly),       big.mark = ","),
                            format(round(gain_ly_pa),    big.mark = ",")),
  `Gain Lwx`           = c("—",
                            format(round(gain_hly_trend), big.mark = ","),
                            format(round(gain_hly),       big.mark = ","),
                            format(round(gain_hly_pa),    big.mark = ","))
) |> knitr::kable()
```

::: {.callout-tip}
## Interpretation
A +2 mMET/week PA increase for the 50–54 female cohort produces a PIF of `r round(pif_female_5054 * 100, 2)`% and a gain of `r format(round(gain_ly_pa), big.mark = ",")` person-years. The trend scenario (`mx_trend = 0.5`) adds `r format(round(gain_ly_trend), big.mark = ",")` person-years from background secular improvement — comparable in magnitude to the PA intervention but operating through a completely different mechanism. In a full analysis both would typically be included simultaneously.
:::

---

## Exercise — extend to multiple cohorts {#sec-exercise}

Try running the full pipeline for **males aged 52** and for **females aged 37**, both with and without the 0.5% trend. What happens to the absolute life-year gains? Why does the trend matter more for younger cohorts?

```{r}
#| label: exercise-multiple-cohorts
#| eval: false

for (sex_i in c("female", "male")) {
  for (age_i in c(37, 52, 62)) {

    age_cat_i <- bogota_mmets %>%
      filter(age == age_i) %>%
      pull(age_cat) %>%
      unique()

    pif_i <- pifs %>%
      filter(sex == sex_i, age_cat == age_cat_i) %>%
      pull(pif_pa)

    lt_ref_i <- RunLifeTable(mslt_general, sex_i, age_i,
                              mx_trend = NA, scenario = 0)
    lt_trend_i <- RunLifeTable(mslt_general, sex_i, age_i,
                                mx_trend = 0.5, scenario = 0)
    lt_pa_i   <- RunLifeTable(mslt_general, sex_i, age_i,
                               mx_trend = NA, scenario = pif_i)
    lt_both_i <- RunLifeTable(mslt_general, sex_i, age_i,
                               mx_trend = 0.5, scenario = pif_i)

    cat(sprintf(
      "%s aged %d | PIF: %.2f%% | Gain PA: %s | Gain trend: %s | Gain both: %s\n",
      sex_i, age_i,
      pif_i * 100,
      format(round(sum(lt_pa_i$Lx)   - sum(lt_ref_i$Lx)),  big.mark = ","),
      format(round(sum(lt_trend_i$Lx) - sum(lt_ref_i$Lx)), big.mark = ","),
      format(round(sum(lt_both_i$Lx)  - sum(lt_ref_i$Lx)), big.mark = ",")
    ))
  }
}
```

---

# Disease Tables — Three-State Model {#sec-disease}

So far we have worked with a **two-state life table**: alive → dead, driven by all-cause mortality. The multi-state life table extends this by adding a **disease layer** for each condition of interest. Each disease follows the three-state process from Barendregt et al. (1998):

$$\text{Healthy} (S_x) \xrightarrow{\text{incidence}} \text{Diseased} (C_x) \xrightarrow{\text{case fatality}} \text{Dead from disease} (D_x)$$

with the constraint $S_x + C_x + D_x = 1{,}000$ at every age.

`RunDisease()` implements this process. It accepts the same `inc_trend` and `cf_trend` scalar parameters as `RunLifeTable()` accepts `mx_trend` — a compounding annual percentage change applied across simulation years:

$$\text{incidence}[t] = \text{incidence}[1] \times \left(1 - \frac{\text{inc\_trend}}{100}\right)^{t-1}$$

## Disease table variables {#sec-dis-vars}

| Variable | Meaning |
|---|---|
| `Sx` | Healthy people at age x (out of 1,000) |
| `Cx` | People with the disease at age x |
| `Dx` | Dead from disease at age x |
| `px` | Prevalence = Cx / (Sx + Cx) |
| `mx` | Disease-specific mortality rate |
| `incidence_disease` | Rate of new disease onset |
| `case_fatality_disease` | Rate of dying given diseased |
| `dw` | Disability weight for this disease |
| `PYx` | Person-years at risk |

## Reference disease tables (no trends) {#sec-dis-ref}

We run reference disease tables for all three Colombia diseases for the same cohort as the life table (females, age 52).

```{r}
#| label: disease-reference

# ── IHD reference ─────────────────────────────────────────────────────────────
dis_ref_ishd <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "ishd",
  inc_trend    = NA,    # static rates
  cf_trend     = NA,
  scenario_inc = 0,
  scenario_cf  = 0
) %>% mutate(label = "Reference")

# ── COPD reference ────────────────────────────────────────────────────────────
dis_ref_copd <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "copd",
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0,
  scenario_cf  = 0
) %>% mutate(label = "Reference")

# ── Type 2 diabetes reference ─────────────────────────────────────────────────
dis_ref_dmt2 <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "dmt2",
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0,
  scenario_cf  = 0
) %>% mutate(label = "Reference")

# Inspect IHD first 8 rows
dis_ref_ishd %>%
  select(age, incidence_disease, case_fatality_disease, Sx, Cx, Dx, px, mx) %>%
  head(8)
```

```{r}
#| label: plot-disease-reference
#| fig-cap: "Disease prevalence (left) and disease-specific mortality rate (right) for IHD, COPD and type 2 diabetes — females aged 52+, Colombia, reference (no trends)."

dis_all_ref <- bind_rows(dis_ref_ishd, dis_ref_copd, dis_ref_dmt2)

pp1 <- ggplot(dis_all_ref, aes(x = age, y = px, colour = disease)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  scale_colour_manual(values = c("ishd" = "#DC2626", "copd" = "#D97706", "dmt2" = "#7C3AED")) +
  labs(title = "Disease prevalence by age", x = "Age", y = "Prevalence (px)", colour = "Disease") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")

pp2 <- ggplot(dis_all_ref, aes(x = age, y = mx, colour = disease)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  scale_colour_manual(values = c("ishd" = "#DC2626", "copd" = "#D97706", "dmt2" = "#7C3AED")) +
  labs(title = "Disease-specific mortality rate", x = "Age", y = "Mortality rate (mx)", colour = "Disease") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")

gridExtra::grid.arrange(pp1, pp2, ncol = 2)
```

## Adding an incidence trend {#sec-dis-trend}

Exactly as `mx_trend` in `RunLifeTable()`, `inc_trend = 0.5` applies a 0.5% annual compounding reduction in incidence across simulation years. The formula is:

$$\text{incidence}[t] = \text{incidence}[1] \times (0.995)^{t-1}$$

```{r}
#| label: disease-trend

dis_trend_ishd <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "ishd",
  inc_trend    = 0.5,   # 0.5% annual decline in incidence, compounding
  cf_trend     = NA,
  scenario_inc = 0,
  scenario_cf  = 0
) %>% mutate(label = "With trend (inc −0.5%/yr)")

# Compare IHD prevalence reference vs trend
dis_ishd_compare <- bind_rows(dis_ref_ishd, dis_trend_ishd)

ggplot(dis_ishd_compare, aes(x = age, y = px, colour = label)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Reference" = "#DC2626",
                                 "With trend (inc −0.5%/yr)" = "#7C3AED")) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  labs(title = "IHD prevalence: reference vs 0.5% annual incidence decline",
       x = "Age", y = "Prevalence (px)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")
```

::: {.callout-note}
## Why prevalence at older ages can be similar despite lower incidence
A declining incidence lowers prevalence at younger ages but the effect partly attenuates at older ages. Fewer new incident cases mean fewer people also dying from the disease — so more people accumulate in the prevalent pool. This is the dynamic captured in Barendregt 1998 Fig. 2 and is a key insight of the proportional MSLT.
:::

## Disease Scenario 1 — 5% reduction in incidence {#sec-dis-sc5}

**Question:** What happens to disease prevalence and disease-specific mortality if incidence falls by 5% permanently at all ages?

```{r}
#| label: disease-scenario-5pct

# Run 5% scenario for all three diseases
dis_sc5_ishd <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "ishd",
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0.05,    # 5% one-off reduction in incidence
  scenario_cf  = 0
) %>% mutate(label = "5% incidence reduction")

dis_sc5_copd <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "copd",
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0.05,
  scenario_cf  = 0
) %>% mutate(label = "5% incidence reduction")

dis_sc5_dmt2 <- RunDisease(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = "dmt2",
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0.05,
  scenario_cf  = 0
) %>% mutate(label = "5% incidence reduction")
```

```{r}
#| label: disease-scenario-5pct-results

# Compare reference vs 5% scenario for each disease
for (dis_name in c("ishd", "copd", "dmt2")) {
  ref <- get(paste0("dis_ref_", dis_name))
  sc5 <- get(paste0("dis_sc5_", dis_name))

  cat(sprintf("\n── %s ──────────────────────────────\n", toupper(dis_name)))
  cat(sprintf("  Cases prevented (sum Cx diff):           %s\n",
              format(round(sum(sc5$Cx) - sum(ref$Cx)), big.mark = ",")))
  cat(sprintf("  Disease deaths prevented (sum Dx diff):  %s\n",
              format(round(sum(sc5$Dx) - sum(ref$Dx)), big.mark = ",")))
  cat(sprintf("  Mean prevalence change:                  %.4f → %.4f\n",
              mean(ref$px), mean(sc5$px)))
}
```

```{r}
#| label: plot-disease-scenario-5pct
#| fig-cap: "IHD, COPD and DMT2 incidence rates: reference vs 5% one-off reduction. The gap between curves is constant across ages (unlike a trend, which accumulates)."

dis_all_sc5 <- bind_rows(
  dis_ref_ishd, dis_sc5_ishd,
  dis_ref_copd, dis_sc5_copd,
  dis_ref_dmt2, dis_sc5_dmt2
)

ggplot(dis_all_sc5, aes(x = age, y = incidence_disease, colour = disease, linetype = label)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  scale_colour_manual(values = c("ishd" = "#DC2626", "copd" = "#D97706", "dmt2" = "#7C3AED")) +
  scale_linetype_manual(values = c("Reference" = "solid", "5% incidence reduction" = "dashed")) +
  labs(title = "Incidence rate: reference vs 5% reduction — IHD, COPD, DMT2",
       x = "Age", y = "Incidence rate", colour = "Disease", linetype = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")
```

::: {.callout-tip}
## Key teaching point — disease scenario vs life table scenario
`scenario_inc = 0.05` in `RunDisease()` reduces incidence by 5% at every age. The full effect on life years only materialises inside `CalculationModel()`, which propagates the disease mortality reduction into the general life table. The disease table alone shows you the epidemiological change; the MSLT shows you the life-years consequence.
:::

---

# Complete MSLT — Physical Activity and Diseases {#sec-mslt}

The full proportional multi-state life table (Barendregt et al. 1998) combines:

1. A **general life table** tracking total survival (from `RunLifeTable()`)
2. **Disease sections** for each modelled condition (from `RunDisease()`)
3. **Feedback**: disease-specific mortality from step 2 modifies total `mx` in step 1

`CalculationModel()` runs all three components simultaneously and returns differences (scenario − reference) in life years, health-adjusted life years, and disease-specific incidence and mortality.

## Disease 2 — PA scenario with disease-specific PIFs {#sec-mslt-pa}

**Question:** What is the gain in life years and health-adjusted life years from a +2 mMET PA increase, modelled through disease-specific incidence reductions for IHD, COPD and type 2 diabetes?

The pathway is:

$$\text{PA distribution} \xrightarrow{\text{dose-response}} \text{RR by disease} \xrightarrow{\text{PIF per disease}} \Delta\text{incidence} \xrightarrow{\text{MSLT}} \Delta\text{life years}$$

### Step 1 — Assign disease-specific RRs {#sec-mslt-rrs}

```{r}
#| label: mslt-assign-rrs

# Disease names as used by the drpa package
diseases_mslt <- c(
  "coronary-heart-disease",
  # "chronic-obstructive-pulmonary-disease", not a physical activity disease
  "diabetes",
  "all-cause-mortality"          # also included for the life table component
)

# Assign RRs for reference and counterfactual PA distributions
bogota_ref  <- assignPA_RRs(bogota_mmets,
                                  diseases = diseases_mslt,
                                  mmet_col = "mmet_ref")
bogota_scen <- assignPA_RRs(bogota_mmets,
                                  diseases = diseases_mslt,
                                  mmet_col = "mmet_scen")

# Quick sanity check — all RRs should be <= 1 (PA is protective)
rr_cols <- grep("^rr_", names(bogota_ref), value = TRUE)
cat("Reference RR ranges:\n")
for (col in rr_cols) {
  cat(sprintf("  %-60s min=%.4f  mean=%.4f\n",
              col, min(bogota_ref[[col]]), mean(bogota_ref[[col]])))
}

cat("Scenario RR ranges:\n")
for (col in rr_cols) {
  cat(sprintf("  %-60s min=%.4f  mean=%.4f\n",
              col, min(bogota_scen[[col]]), mean(bogota_scen[[col]])))
}
```

### Step 2 — Calculate disease-specific PIFs {#sec-mslt-pifs}

```{r}
#| label: mslt-pifs

pif_results_mslt <- calculate_pif(pop_ref  = bogota_ref,
                                   pop_scen = bogota_scen)
pifs_mslt        <- pif_results_mslt$pif_table %>% rename(disease = cause)

cat("PIFs by disease, sex and age group:\n")
print(pifs_mslt  %>% arrange(disease, sex, age_cat), n = Inf)
```

### Step 3 — Extract PIFs for the female 50–54 cohort {#sec-mslt-extract}

```{r}
#| label: mslt-extract-pifs

# Helper: pull one PIF value for the working cohort
get_pif <- function(disease_name, sex_val = "female", age_val = "50-54") {
  pifs_mslt %>%
    filter(disease == disease_name, sex == sex_val, age_cat == age_val) %>%
    pull(pif_pa)
}

pif_ishd <- get_pif("coronary_heart_disease")
pif_dmt2 <- get_pif("diabetes")
pif_acm  <- get_pif("all_cause_mortality")   # for the life table component

cat(sprintf("PIFs for females aged 50-54:\n"))
cat(sprintf("  IHD:               %.4f (%.2f%%)\n", pif_ishd, pif_ishd * 100))
cat(sprintf("  Type 2 diabetes:   %.4f (%.2f%%)\n", pif_dmt2, pif_dmt2 * 100))
cat(sprintf("  All-cause (LT):    %.4f (%.2f%%)\n", pif_acm,  pif_acm  * 100))
```

### Step 4 — Run the complete MSLT {#sec-mslt-run}

```{r}
#| label: mslt-validation

# ── Validation: run with zero scenario — all differences must be zero ─────────
output_test <- CalculationModel(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  mx_trend     = NA,
  in_disease   = c("ishd", "copd", "dmt2"),
  inc_trend    = NA,
  cf_trend     = NA,
  scenario_inc = 0,
  scenario_cf  = 0
)

cat("Validation checks (all must be zero):\n")
cat(sprintf("  sum(Lx_diff):              %.6f\n", sum(output_test$Lx_diff)))
cat(sprintf("  sum(Lwx_diff):             %.6f\n", sum(output_test$Lwx_diff)))
cat(sprintf("  sum(inc_num_diff_ishd):    %.6f\n", sum(output_test$inc_num_diff_ishd)))
cat(sprintf("  sum(inc_num_diff_copd):    %.6f\n", sum(output_test$inc_num_diff_copd)))
cat(sprintf("  sum(inc_num_diff_dmt2):    %.6f\n", sum(output_test$inc_num_diff_dmt2)))
```

```{r}
#| label: mslt-pa-scenario

# ── Scenario A: disease-specific PIFs — disease mortality feeds into general LT
# mx is modified through the disease pathway: incidence ↓ → prevalence ↓ →
# disease-specific mortality ↓ → total mx ↓
output_mslt_dis <- CalculationModel(
  in_data        = mslt_general,
  in_sex         = "female",
  in_mid_age     = 52,
  mx_trend       = NA,
  in_disease     = c("ishd", "copd", "dmt2"),
  inc_trend      = NA,
  cf_trend       = NA,
  scenario       = 0,
  scenario_inc   = c(ishd = pif_ishd, copd = 0, dmt2 = pif_dmt2),
  scenario_cf    = 0,
  use_disease_mx = TRUE   # disease mortality changes propagate to general LT
)

write_csv(output_mslt_dis, "outputs/lecture_5_6/mslt_pa_disease_pifs.csv")

# ── Scenario B: same disease incidence changes, but mx modified via all-cause
# PIF directly — disease mortality feedback switched off.
# Disease incidence still changes (same scenario_inc as above) so case counts
# are produced, but the life-year effect enters via the all-cause PIF on mx
# rather than through the disease-mortality pathway.
output_mslt_acm <- CalculationModel(
  in_data        = mslt_general,
  in_sex         = "female",
  in_mid_age     = 52,
  mx_trend       = NA,
  in_disease     = c("ishd", "copd", "dmt2"),
  inc_trend      = NA,
  cf_trend       = NA,
  scenario       = pif_acm,   # all-cause PIF modifies mx directly
  scenario_inc   = c(ishd = pif_ishd, copd = 0, dmt2 = pif_dmt2),
  scenario_cf    = 0,
  use_disease_mx = FALSE  # disease mortality does NOT feed into general LT
)

write_csv(output_mslt_acm, "outputs/lecture_5_6/mslt_pa_allcause_pif.csv")
```

### Step 5 — Results {#sec-mslt-results}

```{r}
#| label: mslt-results

cat("── Scenario A: disease-specific PIFs → disease mortality in general LT ──\n")
cat(sprintf("  Total life years gained  (sum Lx_diff):  %s\n",
            format(round(sum(output_mslt_dis$Lx_diff)),  big.mark = ",")))
cat(sprintf("  Total HA life yrs gained (sum Lwx_diff): %s\n",
            format(round(sum(output_mslt_dis$Lwx_diff)), big.mark = ",")))
cat(sprintf("  Life expectancy gain (ex_diff at entry): %.1f days\n",
            output_mslt_dis$ex_diff[1] * 365))

cat("\n── Disease cases and deaths prevented (Scenario A) ─────────────────────\n")
cat(sprintf("  IHD cases prevented:   %s\n",
            format(round(abs(sum(output_mslt_dis$inc_num_diff_ishd))), big.mark = ",")))
cat(sprintf("  COPD cases prevented:  %s\n",
            format(round(abs(sum(output_mslt_dis$inc_num_diff_copd))), big.mark = ",")))
cat(sprintf("  DMT2 cases prevented:  %s\n",
            format(round(abs(sum(output_mslt_dis$inc_num_diff_dmt2))), big.mark = ",")))
cat(sprintf("  IHD deaths prevented:  %s\n",
            format(round(abs(sum(output_mslt_dis$mx_num_diff_ishd))),  big.mark = ",")))

cat("\n── Scenario B: disease incidence changes + all-cause PIF on mx ──────────\n")
cat(sprintf("  Total life years gained  (sum Lx_diff):  %s\n",
            format(round(sum(output_mslt_acm$Lx_diff)),  big.mark = ",")))
cat(sprintf("  Total HA life yrs gained (sum Lwx_diff): %s\n",
            format(round(sum(output_mslt_acm$Lwx_diff)), big.mark = ",")))
cat(sprintf("  Life expectancy gain (ex_diff at entry): %.1f days\n",
            output_mslt_acm$ex_diff[1] * 365))
cat("  (Disease case counts are identical to Scenario A — same scenario_inc)\n")
```

```{r}
#| label: plot-mslt-results
#| fig-cap: "Left: life years and health-adjusted life years gained at each age — Scenario A (disease mortality pathway) vs Scenario B (all-cause PIF). Right: disease incidence and mortality changes (identical for both scenarios as they share the same scenario_inc)."

# ── Life years comparison: both MSLT scenarios ────────────────────────────────
ly_compare <- bind_rows(
  output_mslt_dis %>%
    transmute(age, sex,
              measure = "Lx (disease mx pathway)",    value = Lx_diff),
  output_mslt_dis %>%
    transmute(age, sex,
              measure = "Lwx (disease mx pathway)",   value = Lwx_diff),
  output_mslt_acm %>%
    transmute(age, sex,
              measure = "Lx (all-cause PIF)",         value = Lx_diff),
  output_mslt_acm %>%
    transmute(age, sex,
              measure = "Lwx (all-cause PIF)",        value = Lwx_diff)
)

p_ly <- ggplot(ly_compare, aes(x = age, y = value, colour = measure)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  scale_colour_manual(values = c(
    "Lx (disease mx pathway)"  = "#2563EB",
    "Lwx (disease mx pathway)" = "#0D9488",
    "Lx (all-cause PIF)"       = "#DC2626",
    "Lwx (all-cause PIF)"      = "#D97706"
  )) +
  labs(title = "Life years and HALYs: disease pathway vs all-cause PIF",
       x = "Age", y = "Difference (scenario − reference)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")

# ── Disease changes plot — same for both scenarios (shared scenario_inc) ───────
output_graphs <- output_mslt_dis %>%
  select(age, sex, inc_num_diff_ishd, inc_num_diff_copd, inc_num_diff_dmt2,
         mx_num_diff_ishd,  mx_num_diff_copd,  mx_num_diff_dmt2) %>%
  pivot_longer(cols = -c(age, sex), names_to = "measure") %>%
  mutate(measure = recode(measure,
    "inc_num_diff_ishd" = "IHD incidence",
    "inc_num_diff_copd" = "COPD incidence",
    "inc_num_diff_dmt2" = "DMT2 incidence",
    "mx_num_diff_ishd"  = "IHD deaths",
    "mx_num_diff_copd"  = "COPD deaths",
    "mx_num_diff_dmt2"  = "DMT2 deaths"))

p_dis <- ggplot(output_graphs, aes(x = age, y = value, colour = measure)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  labs(title = "Disease incidence and deaths: changes by age\n(identical for both MSLT scenarios)",
       x = "Age", y = "Difference (negative = prevented)", colour = NULL) +
  theme_classic() +
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.text = element_text(size = 11), axis.title = element_text(size = 11),
        legend.position = "bottom")

gridExtra::grid.arrange(p_ly, p_dis, ncol = 2)
```

## Comparing all scenarios {#sec-mslt-compare}

```{r}
#| label: mslt-summary-table
#| echo: false

tibble(
  Approach    = c(
    "1. Simple life table — all-cause PIF",
    "2. MSLT — disease PIFs → disease mx → general LT",
    "3. MSLT — disease PIFs + all-cause PIF on mx"
  ),
  `How mx changes` = c(
    "PIF × all-cause mx directly",
    "Incidence ↓ → disease mx ↓ → total mx ↓",
    "Same incidence changes; mx × (1 − pif_acm)"
  ),
  `use_disease_mx` = c("—", "TRUE", "FALSE"),
  `Gain Lx`  = c(
    format(round(gain_ly_pa),                   big.mark = ","),
    format(round(sum(output_mslt_dis$Lx_diff)), big.mark = ","),
    format(round(sum(output_mslt_acm$Lx_diff)), big.mark = ",")
  ),
  `Gain Lwx` = c(
    format(round(gain_hly_pa),                   big.mark = ","),
    format(round(sum(output_mslt_dis$Lwx_diff)), big.mark = ","),
    format(round(sum(output_mslt_acm$Lwx_diff)), big.mark = ",")
  ),
  `IHD cases prevented` = c(
    "—",
    format(round(abs(sum(output_mslt_dis$inc_num_diff_ishd))), big.mark = ","),
    format(round(abs(sum(output_mslt_acm$inc_num_diff_ishd))), big.mark = ",")
  ),
  `Competing risks` = c("No", "Yes", "Yes")
) |> knitr::kable()
```

::: {.callout-tip}
## Interpreting the three-way comparison
**Approaches 2 and 3 use identical disease incidence changes** (`scenario_inc`) so their disease case counts are the same. The difference is entirely in how the mortality effect reaches the general life table:

- **Approach 2** (`use_disease_mx = TRUE`): the mortality reduction is mediated through disease — incidence falls, fewer people get IHD/DMT2, fewer die from those diseases, total mx falls. The life-year gain is bounded by how much IHD and DMT2 mortality contributes to all-cause mortality.
- **Approach 3** (`use_disease_mx = FALSE`): the same incidence changes happen (same case counts) but the general LT mx is reduced directly by the all-cause PIF. This is equivalent to saying: PA reduces mortality through all its pathways (not just the three modelled diseases), but we want to keep tracking disease cases separately.

If approach 2 and approach 3 give similar life-year totals, it suggests the disease-mortality pathway accounts for most of the all-cause PIF in this population. If approach 3 is substantially larger, there are important PA→mortality pathways beyond the three modelled diseases.
:::

## Exercise — extend to multiple cohorts and diseases {#sec-mslt-exercise}

```{r}
#| label: mslt-exercise
#| eval: false

# Run the full MSLT pipeline for males aged 52 and females aged 37
# Compare: how do disease-specific gains differ by sex and entry age?

for (sex_i in c("female", "male")) {
  for (age_i in c(37, 52)) {

    age_cat_i <- case_when(age_i == 37 ~ "35-39", age_i == 52 ~ "50-54")

    pif_ishd_i <- pifs_mslt %>%
      filter(disease == "coronary_heart_disease",
             sex == sex_i, age_cat == age_cat_i) %>%
      pull(pif_pa)
    pif_dmt2_i <- pifs_mslt %>%
      filter(disease == "diabetes",
             sex == sex_i, age_cat == age_cat_i) %>%
      pull(pif_pa)

    out_i <- CalculationModel(
      in_data      = mslt_general,
      in_sex       = sex_i,
      in_mid_age   = age_i,
      mx_trend     = NA,
      in_disease   = c("ishd", "copd", "dmt2"),
      inc_trend    = NA,
      cf_trend     = NA,
      scenario_inc = c(ishd = pif_ishd_i, copd = 0, dmt2 = pif_dmt2_i),
      scenario_cf  = 0
    )

    cat(sprintf(
      "%s aged %d | Lx gain: %s | Lwx gain: %s | IHD cases: %s\n",
      sex_i, age_i,
      format(round(sum(out_i$Lx_diff)),  big.mark = ","),
      format(round(sum(out_i$Lwx_diff)), big.mark = ","),
      format(round(abs(sum(out_i$inc_num_diff_ishd))), big.mark = ",")
    ))
  }
}
```

---

# Integrated MSLT — CalculationModel() {#sec-calcmodel}

`CalculationModel()` is the engine that ties everything together. It runs five internal steps:

1. **Baseline general life table** — `RunLifeTable()` with `scenario = 0`
2. **Baseline disease life tables** — `RunDisease()` with zero scenarios for every disease
3. **Scenario disease life tables** — `RunDisease()` with disease-specific PIFs or reductions
4. **Aggregate disease changes** — sum Δmx and Δpyld across all diseases
5. **Scenario general life table** — `RunLifeTable()` with modified mx and pyld

The function now accepts two independent mechanisms for modifying mortality:

| Parameter | Mechanism | Enters via |
|---|---|---|
| `scenario_inc` / `scenario_cf` | Disease-mediated (incidence → prevalence → disease mx → total mx) | Disease sections feed into general LT |
| `scenario` | Direct PIF on all-cause mx | Applied on top of disease-mediated changes |

Both can be used simultaneously — they are applied independently at the hazard-rate level.

## Validation — zero scenario {#sec-cm-validate}

Always run with all scenarios set to zero first. Every `_diff` column must be exactly zero — if not, there is a bug in the setup.

```{r}
#| label: cm-validate

output_test <- CalculationModel(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = c("ishd", "copd", "dmt2"),
  mx_trend     = NA,
  inc_trend    = NA,
  cf_trend     = NA,
  scenario     = 0,
  scenario_inc = 0,
  scenario_cf  = 0
)

cat("Validation — all must equal zero:\n")
check_cols <- c("Lx_diff", "Lwx_diff", "ex_diff",
                grep("^inc_num_diff|^mx_num_diff", names(output_test), value = TRUE))
for (col in check_cols) {
  val <- sum(output_test[[col]], na.rm = TRUE)
  ok  <- abs(val) < 1e-6
  cat(sprintf("  %-35s %s  (%.8f)\n", col, if (ok) "PASS" else "FAIL", val))
}
```

## Example 1 — 5% direct incidence reduction across all diseases {#sec-cm-sc1}

This is the disease-level equivalent of the 5% all-cause mortality reduction in the simple life table. `scenario_inc = 0.05` as a scalar applies equally to IHD, COPD and DMT2.

```{r}
#| label: cm-sc1-run

output_sc1 <- CalculationModel(
  in_data      = mslt_general,
  in_sex       = "female",
  in_mid_age   = 52,
  in_disease   = c("ishd", "copd", "dmt2"),
  mx_trend     = NA,
  inc_trend    = NA,
  cf_trend     = NA,
  scenario     = 0,
  scenario_inc = 0.05,   # same 5% reduction for all three diseases
  scenario_cf  = 0
)
```

```{r}
#| label: cm-sc1-results

cat("── Life years and HALYs — 5% uniform incidence reduction ──────────────\n")
cat(sprintf("  Life years gained        (sum Lx_diff):  %s\n",
            format(round(sum(output_sc1$Lx_diff)),  big.mark = ",")))
cat(sprintf("  HA life years gained     (sum Lwx_diff): %s\n",
            format(round(sum(output_sc1$Lwx_diff)), big.mark = ",")))
cat(sprintf("  Life expectancy gain:                    %.1f days\n",
            output_sc1$ex_diff[1] * 365))

cat("\n── Disease cases prevented (negative = fewer cases) ──────────────────\n")
for (dis in c("ishd", "copd", "dmt2")) {
  col <- paste0("inc_num_diff_", dis)
  if (col %in% names(output_sc1)) {
    cat(sprintf("  %-6s cases prevented: %s\n", toupper(dis),
                format(round(abs(sum(output_sc1[[col]], na.rm = TRUE))), big.mark = ",")))
  }
}
```

::: {.callout-note}
## Why MSLT gains differ from the simple life table 5% scenario
The simple life table (`RunLifeTable(scenario = 0.05)`) reduces **all-cause** mx by 5% at every age. The MSLT with `scenario_inc = 0.05` reduces **disease incidence** for the three modelled diseases — which then reduces disease-specific mortality, which then feeds into all-cause mx. Because IHD, COPD and DMT2 together account for only a fraction of all-cause mortality, the total mx reduction is smaller, producing fewer life-year gains. This is the correct and more realistic approach when the mechanism is disease-specific.
:::

## Example 2 — PA intervention with disease-specific PIFs {#sec-cm-sc2}

The two MSLT runs (`output_mslt_dis` and `output_mslt_acm`) were already computed in
[Section @sec-mslt-pa]. Here we summarise the key comparison and define `pif_vector`
for reference in the summary table below.

```{r}
#| label: cm-sc2-named-vector

# pif_copd = 0 because COPD is not in the drpa PA dose-response library
pif_copd   <- 0
pif_vector <- c(ishd = pif_ishd, copd = 0, dmt2 = pif_dmt2)

cat("Disease-specific PIFs for the PA scenario (females, 50-54):\n")
cat(sprintf("  IHD:             %.4f (%.2f%%)\n", pif_ishd, pif_ishd * 100))
cat(sprintf("  COPD:            0.0000 (0.00%%)  — not a PA dose-response disease\n"))
cat(sprintf("  Type 2 diabetes: %.4f (%.2f%%)\n", pif_dmt2, pif_dmt2 * 100))
```

```{r}
#| label: cm-sc2-results

cat("── PA disease PIFs — disease mortality pathway (output_mslt_dis) ────────\n")
cat(sprintf("  Life years gained        (sum Lx_diff):  %s\n",
            format(round(sum(output_mslt_dis$Lx_diff)),  big.mark = ",")))
cat(sprintf("  HA life years gained     (sum Lwx_diff): %s\n",
            format(round(sum(output_mslt_dis$Lwx_diff)), big.mark = ",")))
cat(sprintf("  Life expectancy gain:                    %.1f days\n",
            output_mslt_dis$ex_diff[1] * 365))

cat("\n── PA disease PIFs — all-cause PIF on mx (output_mslt_acm) ─────────────\n")
cat(sprintf("  Life years gained        (sum Lx_diff):  %s\n",
            format(round(sum(output_mslt_acm$Lx_diff)),  big.mark = ",")))
cat(sprintf("  HA life years gained     (sum Lwx_diff): %s\n",
            format(round(sum(output_mslt_acm$Lwx_diff)), big.mark = ",")))
cat(sprintf("  Life expectancy gain:                    %.1f days\n",
            output_mslt_acm$ex_diff[1] * 365))

cat("\n── Disease cases prevented (identical for both — same scenario_inc) ─────\n")
for (dis in c("ishd", "copd", "dmt2")) {
  col_inc <- paste0("inc_num_diff_", dis)
  col_mx  <- paste0("mx_num_diff_",  dis)
  cat(sprintf("  %s — cases prevented: %s  |  deaths prevented: %s\n",
              toupper(dis),
              format(round(abs(sum(output_mslt_dis[[col_inc]], na.rm = TRUE))), big.mark = ","),
              format(round(abs(sum(output_mslt_dis[[col_mx]],  na.rm = TRUE))), big.mark = ",")))
}
```

::: {.callout-tip}
## Interpreting the difference between the two MSLT PA scenarios
Both scenarios use identical `scenario_inc` so they produce identical disease case counts.
The difference in life years reveals how much of the all-cause PIF is explained by the
disease-mortality pathway alone. If `output_mslt_dis` and `output_mslt_acm` give similar
totals, the three modelled diseases account for most of the PA mortality benefit in this
cohort. If `output_mslt_acm` is substantially larger, there are important PA pathways
(mental health, musculoskeletal, other cancers) not captured in IHD, COPD and DMT2.
:::



## Summary comparison across all approaches {#sec-cm-compare}

```{r}
#| label: cm-comparison-table
#| echo: false

tibble(
  Approach = c(
    "Simple LT — direct all-cause PIF",
    "Simple LT — 5% all-cause reduction",
    "MSLT — 5% uniform incidence reduction",
    "MSLT — disease PIFs, disease mx pathway",
    "MSLT — disease PIFs, all-cause PIF on mx"
  ),
  `Scenario parameters` = c(
    "scenario = pif_acm",
    "scenario = 0.05",
    "scenario_inc = 0.05 (all diseases)",
    "scenario_inc = pif_vector; use_disease_mx = TRUE",
    "scenario_inc = pif_vector; scenario = pif_acm; use_disease_mx = FALSE"
  ),
  `Gain Lx` = c(
    format(round(gain_ly_pa),                   big.mark = ","),
    format(round(gain_ly),                       big.mark = ","),
    format(round(sum(output_sc1$Lx_diff)),       big.mark = ","),
    format(round(sum(output_mslt_dis$Lx_diff)),  big.mark = ","),
    format(round(sum(output_mslt_acm$Lx_diff)),  big.mark = ",")
  ),
  `Gain Lwx` = c(
    format(round(gain_hly_pa),                   big.mark = ","),
    format(round(gain_hly),                       big.mark = ","),
    format(round(sum(output_sc1$Lwx_diff)),       big.mark = ","),
    format(round(sum(output_mslt_dis$Lwx_diff)),  big.mark = ","),
    format(round(sum(output_mslt_acm$Lwx_diff)),  big.mark = ",")
  ),
  `Disease cases` = c("not modelled", "not modelled",
                       "yes — all diseases equal",
                       "yes — disease-specific",
                       "yes — disease-specific"),
  `Competing risks` = c("No", "No", "Yes", "Yes", "Yes")
) |> knitr::kable()
```

```{r}
#| label: cm-plot-all-scenarios
#| fig-cap: "Life years gained at each age across all four MSLT scenarios. The simple life table scenarios are shown for reference. Note how the MSLT disease-specific scenarios produce different age profiles of benefit — IHD prevention is concentrated at older ages while diabetes prevention is more uniformly distributed."

# Compile age profiles for plotting
all_scenarios <- bind_rows(
  output_sc1         %>% transmute(age, value = Lx_diff, scenario = "5% uniform incidence"),
  output_mslt_dis    %>% transmute(age, value = Lx_diff, scenario = "PA disease PIFs (disease mx)"),
  output_mslt_acm    %>% transmute(age, value = Lx_diff, scenario = "PA disease PIFs (all-cause PIF)")
)

ggplot(all_scenarios, aes(x = age, y = value, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(limits = c(52, 100), breaks = seq(52, 100, 10), expand = c(0, 0)) +
  scale_colour_manual(values = c(
    "5% uniform incidence"             = "#DC2626",
    "PA disease PIFs (disease mx)"     = "#D97706",
    "PA disease PIFs (all-cause PIF)"  = "#0D9488"
  )) +
  labs(title   = "Life years gained by age — MSLT scenarios compared",
       x       = "Age",
       y       = "Life years gained (Lx_diff)",
       colour  = NULL) +
  theme_classic() +
  theme(plot.title      = element_text(size = 12, hjust = 0.5),
        axis.text       = element_text(size = 11),
        axis.title      = element_text(size = 11),
        legend.position = "bottom")
```

::: {.callout-tip}
## What the comparison teaches
The three MSLT scenarios produce different results because they enter the model at different points in the causal chain:

- **5% uniform incidence** — same proportional reduction for all three diseases regardless of PA dose-response. A useful hypothetical benchmark.
- **PA disease PIFs** — realistic: IHD benefits most from PA, COPD least. The gains are concentrated in the diseases where PA has the strongest dose-response.
- **PA combined** — adds benefits from PA pathways outside the three modelled diseases. Larger total gain but risks double-counting unless the direct `scenario` is carefully specified as the residual.

The simple life table approaches (applying a PIF directly to all-cause mx) produce similar totals to the MSLT approaches but lose all information about which diseases drove the change, how disease prevalence shifted, and how competing risks attenuated the gains.
:::
